### 2016-08-18 Chapters/Ch03-PCMCoupledCluster.tex
Save the day's work. Energy corrections up to fifth order re-checked and typed in

In this Chapter, I will summarize previous and current work on the
formulation of quantum/classical polarizable models in the context of
the highly accurate \acl{CC} model.
This is an ongoing and yet unpublished collaboration with prof.~T.~Daniel Crawford
(Virginia Tech) and Dr.~Andrew C.~Simmonett (NIH)
The first part of this Chapter summarizes existing work, mainly by Cammi
and Caricato, by recasting it in language of variational
quantum/classical polarizable Hamiltonians.
The second part of this Chapter illustrates how the variational language
facilitates the formulation of a \acl{CC}-based \acl{PT} (\acs{CC}-\acs{PT})
including a classical polarizable environment.
I will finally show how the perturbation series can be successfully
exploited to derive iterative and noniterative approximate \acs{CC}
schemes.

@@

    \PCM & \MMPCM^\dagger \\
    \MMPCM & \MM

@@

   \sigma^\mathrm{HF} \\
   \kappa^\mathrm{HF}

@@

   \esp^\mathrm{HF} \\
   \zeta^\mathrm{HF}

@@

    \braket{\mathrm{HF} | \overline{\esp} - \esp^\mathrm{HF} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\esp} - \esp^\mathrm{HF} | \mathrm{HF}} \\
  \braket{\mathrm{HF} | \overline{\zeta} - \zeta^\mathrm{HF} | \mathrm{HF}}

@@

  \zeta^\mathrm{HF}| \mathrm{HF}}

@@

  \frac{1}{2}(\sigma^\mathrm{HF}+\sigma)\PCM(\sigma^\mathrm{HF}+\sigma)
+ (\sigma^\mathrm{HF}+\sigma)(\esp^\mathrm{HF}+\tilde{\esp}(\tamp{},\tbar{})) \\
&+ \frac{1}{2}(\kappa^\mathrm{HF}+\kappa)\MM(\kappa^\mathrm{HF}+\kappa)
+ (\kappa^\mathrm{HF}+\kappa)(\zeta^\mathrm{HF}+\tilde{\zeta}(\tamp{},\tbar{})) \\
&+ (\sigma^\mathrm{HF}+\sigma)\bi{X}(\kappa^\mathrm{HF}+\kappa)

@@

= \left/ \text{Expand and collect} \right/ \\

@@

    \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma^\mathrm{HF}
  + \sigma^\mathrm{HF}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa^\mathrm{HF}
  + \kappa^\mathrm{HF}\zeta^\mathrm{HF}
  + \sigma^\mathrm{HF}\bi{X}\kappa^\mathrm{HF} \\

@@

&+  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma
  + \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa
  + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{HF}\bi{X}\kappa \\
&+  \frac{1}{2}\sigma\PCM\sigma^\mathrm{HF}
  + \sigma\esp^\mathrm{HF}
  + \frac{1}{2}\kappa\MM\kappa^\mathrm{HF}
  + \kappa\zeta^\mathrm{HF}
  + \sigma\bi{X}\kappa^\mathrm{HF} \\

@@

  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma^\mathrm{HF}

@@

  \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa^\mathrm{HF}

@@

  \sigma^\mathrm{HF}\esp^\mathrm{HF}

@@

  \kappa^\mathrm{HF}\zeta^\mathrm{HF}

@@

  \sigma^\mathrm{HF}\bi{X}\kappa^\mathrm{HF}

@@

  \frac{1}{2}\sigma^\mathrm{HF}\esp^\mathrm{HF}

@@

  \frac{1}{2}\kappa^\mathrm{HF}\zeta^\mathrm{HF}

@@

    \frac{1}{2}\sigma^\mathrm{HF}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{HF}\zeta^\mathrm{HF} \\

@@

  \frac{1}{2}\sigma\PCM\sigma^\mathrm{HF}

@@

  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma;\quad
  \frac{1}{2}\kappa\MM\kappa^\mathrm{HF}

@@

  \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa; \quad
  \sigma^\mathrm{HF}\bi{X}\kappa

@@

  \kappa\bi{X}^\dagger\sigma^\mathrm{HF}

@@

  &+ \sigma(\cancelto{0}{\bi{Y}\sigma^\mathrm{HF} + \bi{X}\kappa^\mathrm{HF} + \esp^\mathrm{HF}})
  + \kappa(\cancelto{0}{\bi{W}\kappa^\mathrm{HF} + \bi{X}^\dagger\sigma^\mathrm{HF} + \zeta^\mathrm{HF}}) \\
&+ \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{}) \\

@@

    \frac{1}{2}\sigma^\mathrm{HF}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{HF}\zeta^\mathrm{HF}

@@

&+ \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})

@@

  + \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) +
  \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})

@@

  \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  + \p^{\mathrm{HF}, \dagger}\tilde{\s}(\tamp{},\tbar{})

@@

One can then write the effective Lagrangian as:
\begin{equation}
  \begin{aligned}
  \lag{M}(\tamp{}, \tbar{}, \p) &=
  \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tamp{i} | \overline{H_0} | \mathrm{HF}} \\
  &+
  \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  + \p^{\mathrm{HF}, \dagger}\tilde{\s}(\tamp{},\tbar{})
  + U_\mathrm{pol}^\mathrm{ref}
  \end{aligned}
\end{equation}
In accordance with previous work,\autocite{Cammi2009-gu, Caricato2011-tx} the
Hamiltonian $H_0$ is augmented with the
$\p^{\mathrm{HF},\dagger}\tilde{\s}(\tamp{},\tbar{})$ term
to yield the \acl{PTE} (\acs{PTE}) Hamiltonian -- $H$ -- and the
eventually the \acl{PTED} (\acs{PTED}) Lagrangian:
\begin{equation}\label{eq:pted-cc}
  \lag{M}(\tamp{}, \tbar{}, \p) =
  {}^\mathrm{PTE}\lag{M}
  + \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
\end{equation}
where the \acs{PTE} Lagrangian has been introduced:
\begin{equation}
{}^\mathrm{PTE}\lag{M}
  =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
  + U_\mathrm{pol}^\mathrm{ref}
\end{equation}

@@

We first expand the source terms entering the Lagrangian as follows:
\begin{equation}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \s + \BCHfirst{\s}{T} +
  \BCHsecond{\esp}{T} | \mathrm{HF}},
\end{equation}
and since $\s$ has particle rank $m_\s = 1$ we can conclude that the
following relationships hold:
\begin{subequations}
\begin{align}
  \braket{\mathrm{HF} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad 0 \leq t \leq 1
  \Longleftrightarrow t = 1 \\
  \braket{\mathrm{HF} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad t + u = 1 \\
  \braket{\tbar{i} |
  \s
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i = 1 \\
  \braket{\tbar{i} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i \leq t \leq i + 1 \\
  \braket{\tbar{i} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad t + u = i + 1
\end{align}
\end{subequations}
For a general truncation level we accordingly obtain:
\begin{equation}\label{eq:source-term}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
\end{equation}
Furthermore, for the \acs{CCS} model ($\mathcal{M} = 1$):
\begin{equation}
  \tilde{\s}^1(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \braket{\tbar{1} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \frac{1}{2}\braket{\tbar{1} | \commutator{\commutator{\s}{T_1}}{T_1} | \mathrm{HF}}
\end{equation}
while for the \acs{CCSD} model ($\mathcal{M} = 2$):
\begin{equation}
  \tilde{\s}^2(\tamp{}, \tbar{}) =
  \tilde{\s}^1(\tamp{}, \tbar{})
  + \braket{\tbar{1} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}
\end{equation}
Eventually, within the \acs{CCSDT} model ($\mathcal{M} = 3$) four more
terms are added:
\begin{equation}
  \begin{aligned}
  \tilde{\s}^3(\tamp{}, \tbar{}) &=
  \tilde{\s}^2(\tamp{}, \tbar{})
  + \braket{\tbar{2} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \BCHfirst{\s}{T_3} | \mathrm{HF}}
  + \frac{1}{2}
  \braket{\tbar{3} | \commutator{\commutator{\s}{T_2}}{T_2} |
  \mathrm{HF}}
  + \braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
  \mathrm{HF}}
  \end{aligned}
\end{equation}
We note that:
\begin{itemize}
  \item singles amplitudes assume a unique role when a quantum/classical
    Hamiltonian is introduced, a role that will be more evident in the
    perturbative treatment,
  \item double commutators involving the highest order cluster
    operators can only lead to nonzero terms when $T_1$ is also
    involved and by projection onto the highest order left \acs{CC}
    state, \emph{e.g.}
    $\braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} |
    \mathrm{HF}} \neq 0$
    and
    $\braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
    \mathrm{HF}} \neq 0$
\end{itemize}

@@

Differentiation of the effective Lagrangian in Eq.~\eqref{eq:pted-cc}
with respect to the variational parameters $\tamp{\mu_i}$,
$\tbar{\mu_i}$ and $\p$ yields the \acs{PTED}-\acs{CC}
equations:
\begin{subequations}
  \begin{align}
   \tampEq{\mu_i}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
   + \p^\dagger\braket{\mu_i | \overline{\s} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-amplitudes}\\
   \tbarEq{\mu_i}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} |
    \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}}
    \nonumber \\
    &+ \p^\dagger\lbrace
    \braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_i}} +
    \commutator{\commutator{\s}{T}}{\cluster{\mu_i}}| \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\overline{\s}}{\cluster{\mu_i}} | \mathrm{HF}}
             \rbrace
             = 0 \label{eq:pted-cc-multipliers}\\
    \Omega_\p(\tamp{}, \tbar{}, \p)
    &=
    \V\p + \tilde{\s}(\tamp{}, \tbar{}) = 0
    \label{eq:pted-cc-polarization}
  \end{align}
\end{subequations}
These equations clearly show:
\begin{enumerate}
  \item The $\tilde{\s}(\tamp{}, \tbar{})$ expectation value term in the
    polarization equation couples the amplitudes, multipliers and
    polarization equations.
    This requires a proper macroiteration/microiteration
    self-consistency scheme for its implementation.~\autocite{Cammi2009-gu, Caricato2010-hx}
  \item As a consequence, a single-point \acs{PTED}-\acs{CC} will
    suffer from a $2\times$ prefactor in its computational cost with
    respect to an \emph{in vacuo} \acs{CC} calculation.
  \item Eq.~\eqref{eq:pted-cc-polarization} is \emph{directly} coupled to
    the multipliers equation, Eq.~\eqref{eq:pted-cc-multipliers}, since
    the source terms directly depends on amplitudes and multipliers, but
    only \emph{indirectly} coupled to the amplitudes equation,
    Eq.~\eqref{eq:pted-cc-amplitudes}.
\end{enumerate}

@@

We will now develop a perturbative expansion of the effective \acs{CC}
Lagrangian and derive the \acs{CC}\acs{PT} series when a
quantum/classical polarizable Hamiltonian is used.
The fluctuation potential $\Phi$ is assumed as the perturbation, hence
it will be considered of order 1: $\Phi = O(1)$.
Orders will be counted assuming closed-shell \acs{HF} reference:
we will use square brackets $[i]$ for an $i$-th order contributions.

@@

\paragraph*{Amplitudes}
\begin{equation}
  \Omega_{\mu_i}^{[1]} = \denom{\mu_i}\tamp{\mu_i}^{[1]}
  + \braket{\mu_i | \Phi | \mathrm{HF}}
  + \p^{\dagger\,[1]}\braket{\mu_i | \bar{\s}^{[0]} | \mathrm{HF}}
  + \p^{\dagger\,[0]}\braket{\mu_i | \bar{\s}^{[1]}  | \mathrm{HF}}
  = 0
\end{equation}
And since $\p^{[0]} = 0$ and $\bar{\s}^{[0]} = \s$:
\begin{equation}\label{eq:1st-order-amp}
  \Omega_{\mu_i}^{[1]} = \denom{\mu_i}\tamp{\mu_i}^{[1]}
  + \braket{\mu_i | \Phi | \mathrm{HF}}
  + \p^{\dagger\,[1]}\braket{\mu_i | \s | \mathrm{HF}}
  = 0
\end{equation}
Considering the singles and doubles excitation manifolds:
\begin{subequations}
  \begin{align}
  \Omega_{\mu_1}^{[1]} &= \denom{\mu_1}\tamp{\mu_1}^{[1]}
  + \p^{\dagger\,[1]}\braket{\mu_1 | \s | \mathrm{HF}}
  = 0 \\
  \Omega_{\mu_2}^{[1]} &= \denom{\mu_2}\tamp{\mu_2}^{[1]}
  + \braket{\mu_2 | \Phi | \mathrm{HF}}
  = 0
  \end{align}
\end{subequations}
Thus, despite the closed-shell \acs{HF} reference, the singles
amplitudes will already appear in first order, due to the
quantum/classical coupling.

>>>>

In this Chapter, I will summarize previous and current work on the
formulation of quantum/classical polarizable models in the context of
the highly accurate \acl{CC} model.
This is an ongoing and yet unpublished collaboration with prof.~T.~Daniel Crawford
(Virginia Tech) and Dr.~Andrew C.~Simmonett (NIH)

The first part of this Chapter summarizes existing work, mainly by Cammi
and Caricato, by recasting it in language of variational
quantum/classical polarizable Hamiltonians.

The second part of this Chapter illustrates how the variational language
facilitates the formulation of a \acl{CC}-based \acl{PT} (\acs{CC}-\acs{PT})
including a classical polarizable environment.

I will finally show how the perturbation series can be successfully
exploited to derive iterative and noniterative approximate \acs{CC}
schemes.

@@

    \PCM & \MMPCM \\
    \MMPCM^\dagger & \MM

@@

   \sigma_\mathrm{HF} \\
   \kappa_\mathrm{HF}

@@

   \esp_\mathrm{HF} \\
   \zeta_\mathrm{HF}

@@

    \braket{\mathrm{HF} | \overline{\esp} - \esp_\mathrm{HF} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\esp} - \esp_\mathrm{HF} | \mathrm{HF}} \\
  \braket{\mathrm{HF} | \overline{\zeta} - \zeta_\mathrm{HF} | \mathrm{HF}}

@@

  \zeta_\mathrm{HF}| \mathrm{HF}}

@@

  \frac{1}{2}(\sigma_\mathrm{HF}+\sigma)\PCM(\sigma_\mathrm{HF}+\sigma)
+ (\sigma_\mathrm{HF}+\sigma)(\esp_\mathrm{HF}+\tilde{\esp}(\tamp{},\tbar{})) \\
&+ \frac{1}{2}(\kappa_\mathrm{HF}+\kappa)\MM(\kappa_\mathrm{HF}+\kappa)
+ (\kappa_\mathrm{HF}+\kappa)(\zeta_\mathrm{HF}+\tilde{\zeta}(\tamp{},\tbar{}))
+ (\sigma_\mathrm{HF}+\sigma)\bi{X}(\kappa_\mathrm{HF}+\kappa) \\

@@

&= \left/ \text{Expand and collect} \right/ \\

@@

    \frac{1}{2}\sigma_\mathrm{HF}\PCM\sigma_\mathrm{HF}
  + \sigma_\mathrm{HF}\esp_\mathrm{HF}
  + \frac{1}{2}\kappa_\mathrm{HF}\MM\kappa_\mathrm{HF}
  + \kappa_\mathrm{HF}\zeta_\mathrm{HF}
  + \sigma_\mathrm{HF}\bi{X}\kappa_\mathrm{HF} \\

@@

&+  \frac{1}{2}\sigma_\mathrm{HF}\PCM\sigma
  + \sigma_\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa_\mathrm{HF}\MM\kappa
  + \kappa_\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma_\mathrm{HF}\bi{X}\kappa \\
&+  \frac{1}{2}\sigma\PCM\sigma_\mathrm{HF}
  + \sigma\esp_\mathrm{HF}
  + \frac{1}{2}\kappa\MM\kappa_\mathrm{HF}
  + \kappa\zeta_\mathrm{HF}
  + \sigma\bi{X}\kappa_\mathrm{HF} \\

@@

  \frac{1}{2}\sigma_\mathrm{HF}\PCM\sigma_\mathrm{HF}

@@

  \frac{1}{2}\kappa_\mathrm{HF}\MM\kappa_\mathrm{HF}

@@

  \sigma_\mathrm{HF}\esp_\mathrm{HF}

@@

  \kappa_\mathrm{HF}\zeta_\mathrm{HF}

@@

  \sigma_\mathrm{HF}\bi{X}\kappa_\mathrm{HF}

@@

  \frac{1}{2}\sigma_\mathrm{HF}\esp_\mathrm{HF}

@@

  \frac{1}{2}\kappa_\mathrm{HF}\zeta_\mathrm{HF}

@@

    \frac{1}{2}\sigma_\mathrm{HF}\esp_\mathrm{HF}
  + \frac{1}{2}\kappa_\mathrm{HF}\zeta_\mathrm{HF} \\

@@

  \frac{1}{2}\sigma\PCM\sigma_\mathrm{HF}

@@

  \frac{1}{2}\sigma_\mathrm{HF}\PCM\sigma;\quad
  \frac{1}{2}\kappa\MM\kappa_\mathrm{HF}

@@

  \frac{1}{2}\kappa_\mathrm{HF}\MM\kappa; \quad
  \sigma_\mathrm{HF}\bi{X}\kappa

@@

  \kappa\bi{X}^\dagger\sigma_\mathrm{HF}

@@

  &+ \sigma(\cancelto{0}{\bi{Y}\sigma_\mathrm{HF} + \bi{X}\kappa_\mathrm{HF} + \esp_\mathrm{HF}})
  + \kappa(\cancelto{0}{\bi{W}\kappa_\mathrm{HF} + \bi{X}^\dagger\sigma_\mathrm{HF} + \zeta_\mathrm{HF}}) \\
&+ \sigma_\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) + \kappa_\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{}) \\

@@

    \frac{1}{2}\sigma_\mathrm{HF}\esp_\mathrm{HF}
  + \frac{1}{2}\kappa_\mathrm{HF}\zeta_\mathrm{HF}

@@

&+ \sigma_\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) + \kappa_\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})

@@

  + \sigma_\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) +
  \kappa_\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})

@@

  \frac{1}{2}{}^t\p\V\p + {}^t\p\tilde{\s}(\tamp{},\tbar{})
  + {}^t\p_\mathrm{HF}\tilde{\s}(\tamp{},\tbar{})

@@

One can then write the effective Lagrangian as:
\begin{equation}
  \begin{aligned}
  \lag{M}(\tamp{}, \tbar{}, \p) &=
  \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tamp{i} | \overline{H_0} | \mathrm{HF}} \\
  &+
  \frac{1}{2}{}^t\p\V\p + {}^t\p\tilde{\s}(\tamp{},\tbar{})
  + {}^t\p_\mathrm{HF}\tilde{\s}(\tamp{},\tbar{})
  + U_\mathrm{pol}^\mathrm{ref}
  \end{aligned}
\end{equation}
In accordance with previous work,\autocite{Cammi2009-gu, Caricato2011-tx} the
Hamiltonian $H_0$ is augmented with the
${}^t\p_\mathrm{HF}\tilde{\s}(\tamp{},\tbar{})$ term
to yield the \acl{PTE} (\acs{PTE}) Hamiltonian -- $H$ -- and the
eventually the \acl{PTED} (\acs{PTED}) Lagrangian:
\begin{equation}\label{eq:pted-cc}
  \lag{M}(\tamp{}, \tbar{}, \p) =
  {}^\mathrm{PTE}\lag{M}
  + \frac{1}{2}{}^t\p\V\p + {}^t\p\tilde{\s}(\tamp{},\tbar{})
\end{equation}
where the \acs{PTE} Lagrangian has been introduced:
\begin{equation}
{}^\mathrm{PTE}\lag{M}
  =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
  + U_\mathrm{pol}^\mathrm{ref}
\end{equation}

@@

We first expand the source terms entering the Lagrangian as follows:
\begin{equation}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \s + \BCHfirst{\s}{T} +
  \BCHsecond{\esp}{T} | \mathrm{HF}},
\end{equation}
and since $\s$ has particle rank $m_\s = 1$ we can conclude that the
following relationships hold:
\begin{subequations}
\begin{align}
  \braket{\mathrm{HF} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad 0 \leq t \leq 1
  \Longleftrightarrow t = 1 \\
  \braket{\mathrm{HF} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad t + u = 1
  \label{eq:HF-double-commutator} \\
  \braket{\tbar{i} |
  \s
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i = 1 \\
  \braket{\tbar{i} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i \leq t \leq i + 1 \\
  \braket{\tbar{i} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad t + u = i + 1
\end{align}
\end{subequations}
For a general truncation level we accordingly obtain:
\begin{equation}\label{eq:source-term}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
\end{equation}
By virtue of Eq.~\eqref{eq:HF-double-commutator} the \acs{HF}
expectation value of the double commutator was dropped.

Furthermore, for the \acs{CCS} model ($\mathcal{M} = 1$):
\begin{equation}
  \tilde{\s}^1(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \braket{\tbar{1} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \frac{1}{2}\braket{\tbar{1} | \commutator{\commutator{\s}{T_1}}{T_1} | \mathrm{HF}}
\end{equation}
while for the \acs{CCSD} model ($\mathcal{M} = 2$):
\begin{equation}
  \tilde{\s}^2(\tamp{}, \tbar{}) =
  \tilde{\s}^1(\tamp{}, \tbar{})
  + \braket{\tbar{1} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}
\end{equation}
Eventually, within the \acs{CCSDT} model ($\mathcal{M} = 3$) four more
terms are added:
\begin{equation}
  \begin{aligned}
  \tilde{\s}^3(\tamp{}, \tbar{}) &=
  \tilde{\s}^2(\tamp{}, \tbar{})
  + \braket{\tbar{2} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \BCHfirst{\s}{T_3} | \mathrm{HF}}
  + \frac{1}{2}
  \braket{\tbar{3} | \commutator{\commutator{\s}{T_2}}{T_2} |
  \mathrm{HF}}
  + \braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
  \mathrm{HF}}
  \end{aligned}
\end{equation}
We note that:
\begin{itemize}
  \item singles amplitudes assume a unique role when a quantum/classical
    Hamiltonian is introduced, a role that will be more evident in the
    perturbative treatment,
  \item double commutators involving the highest order cluster
    operators can only lead to nonzero terms when $T_1$ is also
    involved and by projection onto the highest order left \acs{CC}
    state, \emph{e.g.}
    $\braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} |
    \mathrm{HF}} \neq 0$
    and
    $\braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
    \mathrm{HF}} \neq 0$
\end{itemize}

@@

Differentiation of the effective Lagrangian in Eq.~\eqref{eq:pted-cc}
with respect to the variational parameters $\tamp{\mu_i}$,
$\tbar{\mu_i}$ and $\p$ yields the \acs{PTED}-\acs{CC}
equations:
\begin{subequations}
  \begin{align}
   \tampEq{\mu_i}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
   + {}^t\p\braket{\mu_i | \s | \mathrm{HF}}\delta_{\mu_i\mu_1} \nonumber \\
   &+ {}^t\p\braket{\mu_i | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-amplitudes}\\
   \tbarEq{\mu_i}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} |
    \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}}
    \nonumber \\
    &+
    {}^t\p\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_i}} | \mathrm{HF}}\delta_{\mu_i\mu_1} \nonumber \\
    &+
    {}^t\p\sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\s}{\cluster{\mu_i}} | \mathrm{HF}}
    +
    {}^t\p\sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\commutator{\s}{T}}{\cluster{\mu_i}} | \mathrm{HF}}
             = 0 \label{eq:pted-cc-multipliers}\\
    \Omega_\p(\tamp{}, \tbar{}, \p)
    &=
    \V\p + \tilde{\s}(\tamp{}, \tbar{}) = 0
    \label{eq:pted-cc-polarization}
  \end{align}
\end{subequations}
These equations clearly show:
\begin{enumerate}
  \item The $\tilde{\s}(\tamp{}, \tbar{})$ expectation value term in the
    polarization equation couples the amplitudes, multipliers and
    polarization equations.
    This requires a proper macroiteration/microiteration
    self-consistency scheme for its implementation.~\autocite{Cammi2009-gu, Caricato2010-hx}
  \item As a consequence, a single-point \acs{PTED}-\acs{CC} will
    suffer from a $2\times$ prefactor in its computational cost with
    respect to an \emph{in vacuo} \acs{CC} calculation.
  \item Eq.~\eqref{eq:pted-cc-polarization} is \emph{directly} coupled to
    the multipliers equation, Eq.~\eqref{eq:pted-cc-multipliers}, since
    the source terms directly depends on amplitudes and multipliers, but
    only \emph{indirectly} coupled to the amplitudes equation,
    Eq.~\eqref{eq:pted-cc-amplitudes}.
\end{enumerate}

@@

We will now develop a perturbative expansion of the effective \acs{CC}
Lagrangian and derive the \acs{CC}\acs{PT} series when a
quantum/classical polarizable Hamiltonian is used.
The fluctuation potential $\Phi$ is assumed as the perturbation, hence
it will be considered of order 1: $\Phi = O(1)$.
Orders will be counted assuming a closed-shell \acs{HF} reference:
we will use square brackets $[i]$ for an $i$-th order contributions.
We further assume that real wave functions are used.

@@

\paragraph*{Amplitudes}
\begin{equation}\label{eq:1st-order-amp}
  \Omega_{\mu_i}^{[1]} = \denom{\mu_i}\tamp{\mu_i}^{[1]}
  + \braket{\mu_i | \Phi | \mathrm{HF}}
  + {}^t\p^{[1]}\braket{\mu_i | \s | \mathrm{HF}}\delta_{\mu_i\mu_1}
  = 0
\end{equation}
Considering the singles, doubles and triples excitation manifolds:
\begin{subequations}
  \begin{align}
  \Omega_{\mu_1}^{[1]} &= \denom{\mu_1}\tamp{\mu_1}^{[1]}
  + {}^t\p^{[1]}\braket{\mu_1 | \s | \mathrm{HF}}
  = 0 \\
  \Omega_{\mu_2}^{[1]} &= \denom{\mu_2}\tamp{\mu_2}^{[1]}
  + \braket{\mu_2 | \Phi | \mathrm{HF}}
  = 0 \\
  \Omega_{\mu_3}^{[1]} &= \denom{\mu_3}\tamp{\mu_3}^{[1]} = 0
  \end{align}
\end{subequations}
Thus, despite the closed-shell \acs{HF} reference, the singles
amplitudes will already appear in first order, due to the
quantum/classical coupling.
